---
title: "Lecture 2: Data Structures"
author: "PSYC417"
format: docx
editor: visual
---

# Data Structures Workbook

## Data structures

R classifies objects different ways, which impacts the operations you
can perform and what you can store inside.

### **Vectors**

- Unidimensional and can only contain a single data type. Created with
  the c() command.

- You can check how long a vector is using length()

```{r}
vec <- c("a","b","c","d");vec
length(vec)
```

### **Matrices**

- Two dimensional and can only contain a single data type. Created with
  the matrix() command.

- You can check how many elements are in a matrix using length(). You
  can check how many rows there are using nrow() and you can check how
  many columns there are using ncol().

- byrow = FALSE is the default in matrix(), but you can change that to
  true to fill in values by row instead of by column.

```{r}
mat <- matrix(c(1:9), nrow=3, ncol=3);mat
mat2 <- matrix(c(1:9), nrow = 3, ncol = 3, byrow = TRUE);mat2

length(mat)
nrow(mat)
ncol(mat)
```

### **Data frames**

- Two dimensional and can contain multiple data types. R's primary means
  of data storage and akin to a spreadsheet. Created with the
  data.frame() command.

- You can check the number of rows using nrow() or the number of columns
  using ncol() OR length()

```{r}
df <- data.frame(x = rep(TRUE, 4), y = c(1:4))
length(df)
nrow(df)
ncol(df)
```

### **Lists**

- Multidimensional and capable of storing other objects. Created with
  the list() command.

- You can check how many objects are in a list using length()

```{r}
lis <- list(vec = vec, mat = mat, df = df)
length(lis)
```

### Checking/Converting Structures

You can check the data structure if you're unsure about it using class()
or is.datastructure() syntax.

```{r}
is.vector(vec); class(vec)
is.matrix(mat); class(mat)
is.data.frame(df); class(df)
is.list(lis); class(lis)
```

You can convert some objects to others using as.vector(), as.matrix(),
etc.

```{r}
new_vec <- as.vector(mat);print(new_vec)
```

### Practice 1

Create a different chunk for each of these practice exercises (for
practice).

1.  Build each type of data structure. Try to vary the data types if you
    can. (Don't make every object consist of numbers and only numbers.)

2.  Convert each data structure to a different data structure. Print or
    view the new structures. Which ones DON'T work well?

3.  Double that numbers that exist in relevant vectors, matrices, or
    data frame columns. (Create ones with numbers if you must.)

## Indexing and Subsetting

You can extract values from objects using a variety of operators in R.
The base syntax is to type the object name followed by \[\]. Sometimes
you can use a \$ and other times you can use \[\[\]\].

- R uses 1-based indexing so you just put the position of the element
  you want.

```{r}
# These are the objects from the slides
v <- c("a","b","c","d","e")
m <- matrix(c(1:9), nrow=3,ncol=3)
df <- data.frame(movie=c("Avengers 1","Avengers 2", "Avengers 3 Part 1", "Avengers 3 Part 2"), ratings = c(8,7,10,7),bo_ranking = c(1:4))
tot <- list(vec = v, mat = m, datframe = df)
```

**Vectors**: Put a single number in the brackets to extract the specific
element. You can only use brackets.

```{r}
v[3] # gives us the second value in vec
v[1:3] # gives us the first three values in vec
v[6] = 100 # creates a sixth element in v
```

**Matrices**: Two-dimensional. You need to account for BOTH dimensions
in the bracket index. EXAMPLE\[ROW, COLUMN\]. (Matrices can have named
dimensions, but we rarely use them.) Leaving a dimension blank will give
you all the elements in that dimension.

```{r}
m[2,3] # gives us the value in the second row and the third column
m[2,] # gives us the entire second row
m[,3] # gives us the entire third column
```

**Data Frame**: Two-dimesional. You need to account for BOTH dimensions
in the bracket index. This will give you a vector.

```{r}
df[,1] # entire first column
df[1,1] # first value in the first column
```

You can also use the dollar sign \$ to refer to a dimension by name or
use the name in brackets. This will give you a vector.

```{r}
df$movie
df[,"movie"]
is.vector(df$movie)
```

Putting JUST the name or position gives you a one column data frame.

```{r}
df["movie"]
df[1]
is.data.frame(df[1])
```

Lists: Can extract elements by name using \$ or by position using the
object position in \[\[\]\]. The object will retain its structure.

```{r}
tot[[1]]
tot$vec
```

You can SUBSET values from objects using \[\] or the subset() command.

```{r}
v[v=="c"]

subset(df, ratings > 7)
df[df$ratings > 7, ]

subset(df, ratings == min(df$ratings) | ratings == max(df$ratings))
```

### Practice 2

1.  Add a new data frame column to your existing data frame that
    consists of boolean values. Name it "BOO."

2.  Store every object/structure you've created in a list called
    "complete." Subset the 3rd element of that list and assign it to an
    object called THIRD.

3.  Use the ratings and greetings vectors in the code chunk below to
    complete the following exercises:

    a.) Find how many "bad ratings" (a score of two or less) are in the
    ratings vector. Do this in one line of code.

    b.) Find how many "extreme ratings" there are in one line of code.
    (Hint: You may need to use an OR operator.)

    c.) Type greetings=="hey" and see what happens. Can you then find
    (in one line of code) whether more people say 'hey' than 'hi'?
    (Hint: This will require boolean operators!)

```{r}
ratings <- c(1,3,5,3,3,2,5,5,4,4,3,4,5)
greetings <- c("hey","hi","hey","yo","hello","sup","hey","hi")
```

## Merging

We can combine objects together using various commands.

- **c()** will concatenate objects of similar structures together.
  (e.g., stack vectors on top of each other).

```{r}
v1_5 <- c(1:5);v6_10 <- c(6:10)
c(v1_5, v6_10)
```

- **cbind()** will combine similar-lengthed vectors treating each as a
  column or add columns to similarly-lengthed data frame or matrices (or
  you can add a vector as a column).

```{r}
v1_5 <- c(1:5);v6_10 <- c(6:10)
cbind(v1_5, v6_10)
```

- **rbind()** will combine similar-lengthed vectors treating each as a
  row or add rows to similarly-lengthed AND NAMED data frame or matrices
  (or you can add a vector as a row).

```{r}
v1_5 <- c(1:5);v6_10 <- c(6:10)
rbind(v1_5, v6_10)
```

- **merge()** will merge two data frames together. You put the name of
  the first data frame, followed by the name of the second data frame,
  followed by the column you're joining by.

```{r}
evals=data.frame(ID=c(1:4), Evals=c(8,7,5,5), Like = c(7,7,3,3))
student = data.frame(ID=c(1:4),Major=c(rep("psyc",2), "soc", "cs"), Year=c("junior", "senior","senior","first-year"))
stud_dat <- merge(evals,student,by="ID")
stud_dat
```

### Practice 3

1.  Create a one column data frame called "score", fill it with numbers,
    and merge it with stud_dat.

2.  Extract ID, Evals, Like, and Score from stud_dat and concatenate
    them into a single vector called "numbers".

```{r}

```

## The end

That's it! Great job! :)
