---
title: "Lecture 1: Data Types in R Code"
author: "PSYC417"
format: html
editor: visual
---

# Data Types Workbook

# What is R?

-   R is a free, open-source statistical programming language. You can program in R in environments in RStudio (like you're doing now).

-   Knowing R well will benefit you if you're searching for a career in research or any position in graduate school.

## Some basics

-   You assign values to objects using = or \<- . The names you assign should be informative, unique, brief, not use spaces, and start with a character.

-   You use print(object_name) or simply type the object's name to retrieve its contents.

-   You can put multiple commands in the same line by using semi-colons.

-   You can leave comments for readers using the \# symbol. Nothing after that symbol will be interpreted by R.

## Practice

a.) Create 2 objects with a number inside of them, one using each assignment operator. Print each object on the same line you create it and leave a comment above each line stating what you are doing.

```{r}

```

b.) Add the two OBJECTS together by their name. Assign this value to a third object. Leave another comment above the line stating what you are doing.

```{r}

```

## Data types

-   **Numeric** - Whole numbers and integers

-   **Integer** - Whole numbers only, more efficient than numeric

-   **Character** (aka string) - Text data

-   **Logical** (aka Boolean) - TRUE/FALSE, usually in an evaluation of a statement

-   **Factor** - Ordered categorical data

You can check the data type if you're unsure using the is.datatype() syntax:

-   is.numeric(), is.integer(), is.character(), is.factor(), and is.logical()

You can also convert data to other types so long as it's appropriate using the as.datatype() syntax:

-   as.numeric(), as.integer(), as.character(), as.factor(), and as.logical()

```{r}
num1 <- 1; class(num1);
num2 <- 5.62234; class(num2)
int <- 4L; class(int)
char1 <- "A"; class(char1)
char2 <- "The quick brown fox jumped over the lazy dog."; class(char2)
boo1 <- TRUE; class(boo1)
boo2 <- 3==3; class(boo2)
fact <- factor(c("Assistant","Associate","Full"),levels=c("Assistant","Associate","Full"));class(year) # error? how to fix?
```

Try fixing these data types below

```{r}
# I want the below to be logical
boohoo = "false"
# check that it's logical

# I want the below to be numeric
intish = 4.6L
# check that it's numeric

# I wawnt the below to be an ordered factor with the levels first, second, third, last. For some reason it's in alphabetical order!
facty = factor(c("First","Second","Third","Last"))
# check that it's a factor

# I want the below to be an integer
numy = "8"
# check that it's an integer


# I want the below to be character
chard = chard
#check that it's character

```

## Practice exercises

Create a different chunk for each of these practice exercises (for practice).

1.  Create an object called "four" with the value 3 and an object "three" with the value 4 and check: 1.) whether three \> four, (2.) whether 3 \> 4, and (3.) whether three == 4.
2.  Check whether your two favorite Ryans in Hollywood are equal (e.g., "Ryan Reynolds" == "Ryan Gosling"). How is this comparison happening? What about their first names?
3.  Create a variable called X and assign it a value of 10. Then, on the same line, check if the square root of X is greater than the log of X.
4.  Create a variable called Y and assign it to a value of 709. ON the next line, create a variable called Z and assign it a value of 308. Check whether the exponential of Y is greater than or equal to 10 to the power of Z.
5.  Copy the code from the problem above and increase Y and Z by 1. What changed (besides the numbers being 710 and 309)?
6.  Print both Y and Z. What are they equal to?
7.  Create a formula that calculates your grade in a class based on three categories: Exams as 50% of your grade, Homework as 25%, and Attendance as 25%. Create three variables called exam_avg, hw_avg, and attend, and assign them values of your choice. Make sure the grades make sense!

## The end

That's it! Great job! :)
