---
title: "SOLUTIONS Lecture 5: Functions"
author: "PSYC417"
format: docx
editor: visual
---

# Functions Workbook

## Recap

Conditionals (*if*, *else if*, *else*, and *ifelse* statements) and
Loops (*for*, *while*) have a similar layout, with an initiating word
(e.g., *for*), followed by a specification in parentheses, followed by
braces to enclose the body of the loop or conditional. These are two key
aspects to algorithms and thinking like a programmer. The third,
functions, will follow the same structure, both in syntax and purpose.

## Functions

You initiate a function with the key word *function*, specify
'arguments' in parentheses that modify the function output, and put code
within curly braces. One key difference from loops and conditionals: You
should name a function since functions are objects in R. With functions,
you **return** an object at the end of the function code.

Function are great for a number of reasons. For instance:

1)  We improve **readability** and minimize error. Imagine if you had to
    do calculate the mean, standard deviation, etc. by hand every time!

2)  You improve **efficiency**, because you can write functions with
    many additional arguments/options (preventing the need for other
    functions).

3)  They are a **tool/product** that you can make for other people (or
    yourself).

An example of a function is below. It will convert whatever vector the
user enters for the argument x to a numeric data type (within reason):

```{r}
make_numeric<- function(x){
  num_x <- as.numeric(x)
  return(num_x)
}
bad_num <- c("1","2","3")

good_num <- make_numeric(bad_num)

bad_num; good_num
```

### **Availability of Variables**

- Functions run locally. That means that variables defined in functions
  are only available while the function is running. This is a good
  thing. Imagine if your object kept getting overwritten because you
  used a function that had an object with the same name as yours.

- For example, the data [**x**]{.underline} (provided by the user) and
  [num_x]{.underline} (created in the function) only exist while the
  make_numeric() function is running.

- These values are created from scratch each time the function is
  called.

```{r}
demonstrate <- function(x){
  new_data <- x*2
  return(x*2)
}
demonstrate(5);print(new_data)
```

### Practice 1

1.) Let's convert code for calculating the maximum into a function. Call
the function **vmax** and give it one argument: data.

```{r}
vmax <- function(data){ # start function
  temp_max <- -Inf 
  for (i in 1:length(data)) { # start for loop
    if (temp_max < data[i]) { # start if
      temp_max <- data[i]
    } # end if statement
  } # end for loop
  return(temp_max)
} # end function
```

Now test it with different vectors you create!

```{r}
# test on x 
x <- c(10,40,3,12,18); vmax(x)
# create an object y and test it on that 
y <- seq(1000, 2000, by= 123); vmax(y)
```

2.) Create two different functions for the median. First, create
**sort_median()** that calculates the median manually. Then, create
**R_median()** that uses the **median()** function in base R. A few
notes:

- You will need the **sort()** function, which sorts a vector from least
  to greatest.

- You may need the **floor()** command, which returns a truncated whole
  number (i.e., the input rounded down).

- You may need the **ceiling()** command, which returns the input
  rounded up.

- You may also use the **%%** (modulo) operator, which tells you the
  remainder after division.

```{r}
# version 1 - sort_median()
sort_median = function(data){ # open function
  n = length(data)
  s_dat = sort(data)
  if(floor(n/2) != n/2){
    med = s_dat[ceiling(n/2)]
  } else if(floor(n/2) == n/2){ # could use else instead, same result
    med = mean(s_dat[(n/2):(n/2+1)])
  }
  return(med)
} # close function

# version 2 - R_median() 
R_median <- function(data){ # open function
  return(median(data))
} # close function 
```

Test the functions with vectors of different sizes. End by testing with
c(1:100000000)!

```{r}
iterate_mean(1:100);vectorized_mean(1:100)
iterate_mean(1:100000);vectorized_mean(1:100000)
iterate_mean(1:100000000);vectorized_mean(1:100000000)
```

## Multiple Arguments

You can add multiple arguments that further customize the function to
your or the user's needs.

```{r}
vmax2 <- function(data, n){
   temp_max <- -Inf
   for(i in 1:n){
      if(temp_max < data[i]){
         temp_max = data[i]
      } # end if statement
   }# end for loop
   return(temp_max)
} # end function

# consider the following attempts at running a function
vmax2(c(1:10)) # wwrong because missing argument
vmax2(10, c(1:10)) # fine
vmax2(c(1:10), 10) # wrong because arugments out of order and not named
```

### Default Arguments

When you have multiple arguments, you can set a default value that will
run unless the user specifies something else.

```{r}
vmax2 <- function(data, n=length(data)){
   temp_max <- -Inf
   for(i in 1:n){
      if(temp_max < data[i]){
         temp_max = data[i]
      } # end if statement
   }# end for loop
   return(temp_max)
} # end function

vmax2(c(1:10)) # now works because it automatically defaults to the length of the vector
```

### Some important notes:

- Make sure you're designing functions to fit the objects you want to
  work with. These functions will not work on data frames as they
  currently are written.

- R reads arguments by position, so if you don't address each argument
  by name, you better know the order the arguments are in.

- It is usually wise-especially when creating functions that others
  use-to build in safeguards that prevent a function from running when
  an argument is misspecified. R's messages aren't always helpful.

- You can add stop() to generate an informative error message of your
  choice. This is a legitimate error message that will prevent your
  document from rendering if present.

### Practice 2

1.) Let's combine our functions from earlier:

- Make a function with three arguments: median_choice(data, type,
  expression).

- Let **type** take on "sort" or "baseR" that trigger the functions
  sort_median() or R_median(), respectively.

- Let expression be "Yes" or "No" and paste an expression, "Your mean is
  XX", if it is set to "Yes". Otherwise, just have it return the number.

```{r}
median_choice <- function(data, type, expression){ # start function) 
  if(type == "sort"){
    med <- sort_median(data)
  } else if(type == "baseR"){
    med <- R_median(data)
  }	
  if(expression == "No"){
    return(med)
  } else if(expression == "Yes"){
    return(paste("Your median is", med))
  }
} # end function 
```

2.) Create a function called "countdown" that counts down (i.e., prints
numbers) from some value, n, to another value, stop. Make the default
value of stop 0.

```{r}
countdown <- function(n, stop = 0){
  for(i in 1:((n-stop)+1)){
    print(n)
    n <- n-1
  }
}
countdown(10,3)
```

A different approach using recursion:

```{r}
countdown2 <- function(n, stop = 0){
  print(n)
  if(n > stop){
    countdown2(n-1, stop)
  }
}

countdown2(10,3)
```

## Special Features with Functions

We've learned a lot about building functions, but we've only just begun.
Writing a function requires you to think about how people will use it,
both correctly and incorrectly. For example: We will want to use
conditional statements to validate arguments are specified correctly,
lest our function return the wrong number. Remember: This is a product
for others. You want everything to be 100% correct and let people know
when they're using it wrong.

```{r}
vmax3 <- function(data, n = length(data)){
   if(class(data)=="numeric" | class(data)=="integer"){
      temp_max <- -Inf
      for(i in 1:n){
         if(temp_max < data[i]){
            temp_max = data[i]
         } # end if statement
      } # end loop
      return(temp_max)
   }else{ 
      stop("Data must be numeric or integer")
   } # end conditional
} # end function
```

## Bonus: Some examples of functions in base R

We can expand our access to functions by downloading packages, but
sometimes what we need isn't covered in either case. Thankfully, we can
now build our own.

### mean(), median(), minimum(), maximum(), range() and sum()

```{r}
v <- c(40,18,60)
mean(v)
median(v)
min(v)
max(v)
range(v)
sum(v)
prod(v)
```

The efficacy of those functions depends on the data type and structure,
so become familiar with how those work.

### sort()

Arranges a vector in ascending or descending order. Similar to the
"arrange" pipe in dplyr.

```{r}
v <- c(18,75,20,33,4,56,12)
#the default will sort in ascending order
sort(v)
#to sort in descending order, you have to specify "decreasing=TRUE" as an argument
sort(v, decreasing=TRUE)
```

With character vectors, it will sort in alphabetical order, given that
this usually the order of the collating sequence.

```{r}
s <- c("e","r","a","g","v","d")
sort(s)

#the decreasing argument also works with characters
sort(s, decreasing=TRUE)
```

It also works when combining characters and numbers but it will convert
the numbers into string and do weird things with those given the
collating sequence (remember "10" \< "2" when specified as a string).

```{r}
sort(c(s,v))
```

### nchar()

Counts the number of character in a character object.

```{r}
#in a word/single string
nchar("banana")
```

```{r}
#in a vector
nchar(s)
```

### exists()

Checks if a variable is defined in your environment. Careful, you have
to specify the variable name as a character string for the input.

```{r}
exists("v")
exists("s")
exists("vmax3")
exists(vmax2) #this won't work
```

## The end

That's it! Great job! :)
