---
title: "Lecture 1: Data Types in R Code"
author: "PSYC417"
format: docx
editor: visual
---

# Data Types Workbook

## What is R?

- R is a free, open-source statistical programming language. You can
  program in R in environments in RStudio (like you're doing now).

- Knowing R well will benefit you if you're searching for a career in
  research or any position in graduate school.

## Some basics

- You assign values to objects using = or \<- . The names you assign
  should be informative, unique, brief, not use spaces, and start with a
  character.

- You use print(object_name) or simply type the object's name to
  retrieve its contents.

- You can put multiple commands in the same line by using semi-colons.

- You can leave comments for readers using the \# symbol. Nothing after
  that symbol will be interpreted by R. You will be expected to write
  neat, well-commented code this semester.

## Practice 1 

a.) Create 2 objects with a number inside of them, one using each
assignment operator.

```{r}
six = 6
seven <- 7
```

b.) Print each object and leave a comment above each line stating what
you are doing.

```{r}
# print the object six
print(six)
# print the object seven
print(seven)
```

c.) Add the two OBJECTS together by their name. Assign this value to a
third object. Leave another comment above the line stating what you are
doing.

```{r}
# adding the two objects above together
sixseven = six + seven; print(sixseven)
```

## Data types

There are various data types in R. Knowing them (and their limits) is
critical to avoiding unnecessary errors.

### **Numeric**

**Numeric** dataconsist of whole numbers and integers. 5.23, 6.0, 7, and
-0.25 are all examples of numeric data.

### **Integer**

**Integer** data consist only of whole numbers denoted by L after.
Integer data are more "efficient" than numeric, as they do not store
trailing zeroes at the end of a number. 1L, -10L, and 100L are all
examples of integer data.

### **Character** (aka string)

**Character** data consist of text surrounded by quotes. The quotes can
be single or double, so long as they are the same for a given character
data. "A", "dog went to the store", and '8' are all examples of chracter
data.

### **Logical** (aka Boolean)

**Logical** data consist of TRUE and FALSE values (styled exactly as
seen). Whether literally TRUE or FALSE, or a statement with a logical
operator, such as 3 \> 2 (which evaluates to TRUE), logical data will
have an important role once we get to conditional statements. Examples
of logical operators are below:

| Logical Operator | Meaning                                |
|------------------|----------------------------------------|
| \>               | Greater than                           |
| \>=              | Greater than or equal to               |
| \<               | Less than                              |
| \<=              | Less than or equal to                  |
| ==               | Is equal to? (Must be TWO =)           |
| !=               | Not equal to                           |
| &                | AND (to tie together multiple clauses) |
| \|               | OR (to tie together multiple clauses)  |

### **Factor**

**Factor** data consist of nominal and ordinal data types. Sometimes
considered a "data structure" (we'll learn about those soon), we
interact with them in a data type capacity (especially for analyses with
categorical variables and data visualization). An example of factor data
is below.

```{r}
school_years = c("junior", "first-year", "junior", "sophomore", "senior", "junior", "junior", "sophomore", "first-year", "senior+", "senior")

# Be careful…R will alphabetize/order factors if you do not specify.

school_years_ordered = factor(school_years, levels = c("first-year", "sophomore", "junior", "senior", "senior+"), ordered=TRUE) # Ordered = TRUE is for ordinal data
# view the data 
school_years_ordered # you will see less than signs between levels

school_years_unordered = factor(school_years, levels = c("first-year", "sophomore", "junior", "senior", "senior+"), ordered=FALSE) 
# view the data
school_years_unordered # you will only see levels listed
```

### Checking Data Types

You can check the data type if you're unsure using the is.type() syntax:

- is.numeric(), is.integer(), is.character(), is.factor(), and
  is.logical() will return a logical TRUE/FALSE depending on the data
  you enter.
- class() will report the data type back to you.

```{r}
# numeric data
num1 <- 1; is.numeric(num1);
num2 <- 5.62234; class(num2)
# integer data
int1 <- 4L; class(int1)
int2 <- 0L; is.integer(int2)
# character data 
char1 <- "A"; class(char1)
char2 <- "The quick brown fox jumped over the lazy dog."; is.character(char2)
# logical data
logi1 <- TRUE; class(logi1)
logi2 <- 3==3; is.logical(logi2)
# factor data
fact <- factor(c("Full", "Associate", "Assistant", "Assistant","Associate","Full","Associate","Full"),levels=c("Assistant","Associate","Full"));class(fact)
```

### Changing Data Types

You can also convert data to other types so long as it is appropriate
using the as.type() syntax format:

- as.numeric(), as.integer(), as.character(), factor(), and as.logical()
  will make sensible conversions to the specified datatype.

```{r}
# chracter data 
as.character(TRUE)
as.character(8)
as.character(0L)
# numeric data 
as.numeric("1")
as.numeric(0L)
as.numeric(FALSE)
# integer data
as.integer(4.23)
as.integer("123")
as.integer(TRUE)
# factor data
places = c("First", "Second", "Last")
factor(places, levels = c("First", "Second", "Last"), ordered = TRUE)
```

### Practice 2

Create a different chunk for each of these practice exercises (for
practice).

1.  Create an object called "four" with the value 3 and an object
    "three" with the value 4 and check: 1.) whether three \> four, (2.)
    whether 3 \> 4, and (3.) whether three == 4.

```{r}
# number 1 
four = 3; three = 4
three > four
3 > 4
three == 4
```

1.  Check whether your two favorite Ryans in Hollywood are equal (e.g.,
    "Ryan Reynolds" == "Ryan Gosling"). How is this comparison
    happening? What about their first names?

```{r}
# number 2 
"Ryan Reynolds" == "Ryan Gosling"
"Ryan Reynolds" > "Ryan Gosling"
# comparisons of text will look at alphabetical order, with later positions being "greater than" earlier positions
"Ryan" ==  "Ryan" # equal as expected
```

1.  Create a variable called X and assign it a value of 10. Then, on the
    same line, check if the square root of X is greater than the log of
    X.

```{r}
# number 3
x = 10; sqrt(x) > log(x)
```

1.  Create a variable called Y and assign it to a value of 709. ON the
    next line, create a variable called Z and assign it a value of 308.
    Check whether the exponential of Y (using the exp() command) is
    greater than or equal to 10 to the power of Z.

```{r}
# number 4
y = 709; z = 308
exp(y) >= 10 ^ z
```

1.  Copy the code from the problem above and increase Y and Z by 1. What
    changed (besides the numbers being 710 and 309)?

```{r}
# number 5 
exp(y + 1) >= 10 ^ (z+1)
```

1.  Print the modified values of Y and Z from question 5. What are they
    equal to?

```{r}
# number 6 
print(exp(710)); print(10^(309))
```

1.  Create a formula that calculates your grade in a class based on
    three categories: Exams as 50% of your grade, Homework as 25%, and
    Attendance as 25%. Create three variables called exam_avg, hw_avg,
    and attend, and assign them values of your choice. Make sure the
    grades make sense!

```{r}
# number 7 
exam_avg = 75; hw_avg = 82; attend = 100
grade = .5*exam_avg + .25*hw_avg + .25*attend
print(grade)
```

1.  Fix the following data types using the as.type() notation when
    possible, not by fixing the code manually.

```{r}
# I want the below to be logical
boohoo = "false"; boohoo = as.logical(boohoo)
# check that it's logical
is.logical(boohoo)

# I want the below to be numeric
intish = 4.6L
# check that it's numeric
is.numeric(intish) # it already is, however we could also convert it (though you shouldn't)
intish = as.numeric(intish); is.numeric(intish)

# I wawnt the below to be an ordered factor with the levels first, second, third, last. For some reason it sometimes ends up in alphabetical order!
facty = c("First","Second","Third","Last")
# check that it's a factor
facty = factor(facty, levels = c("First", "Second","Third","Last"))
print(facty)
is.factor(facty)

# I want the below to be an integer
numy = "8"
# check that it's an integer
numy = as.integer(numy); class(numy)

# I want the below to be character
twelve = 12
#check that it's character
twelve = as.character(twelve); class(twelve)
```

## The end

That's it! Great job! :)
