---
title: "Lecture 3: Conditionals Workbook"
author: "PSYC417"
format: docx
editor: visual
---

# Conditionals Workbook

## Quick review from last time

Data structure of data frame subsets:

-   **Vector** if indexing a single column using \$ or \[ , \].

-   **Data frame** if index using \[\] with no comma.

-   **Data frame** if indexing multiple columns.

```{r}
# Let's create a data frame with IDs, major, and year as the variables
df <- data.frame(id=1:4, major=c("psych","psych","soc","cs"),
                 year=c("junior","senior","senior","freshman"))
df
```

Indexing in action:

```{r}
# if using the $ method for selecting the column
is.vector(df$major)

# if indexing the column using brackets
is.vector(df[,"major"])

# single brackets, by name
is.vector(df["major"]) # false
is.data.frame(df["major"]) # true

# single brackets, by position
is.vector(df[2]) # false 
is.data.frame(df[2]) # true

# multiple columns
is.vector(df[, 2:3]) # false
is.data.frame(df[, 2:3]) # true
```

Quick illustration of the **summary()** function:

-   Produces five-number summary of all numeric data (and descriptives/data type of non-numeric types).

-   Counts NAs for each variable.

```{r}
summary(df)
```

## What's an algorithm?

-   An exact specification of the steps you need in order to solve a problem.
-   In designing an algorithm, we need to take into account two dimensions:
    -   **Computational complexity**: How many 'computations' will it take before I get an answer?
    -   **Memory complexity**: How much storage do I need (for intermediate values)?
-   Multiple algorithms can often be used to solve the same problem (as long as it can be solved).

## Building blocks of algorithms

-   **Conditional statements**: Run a block of code only if a particular condition or set of conditions are met. These must be written as logical statements.
    -   If TRUE, run the block of code.
    -   If FALSE, ignore the block of code.
-   **Loops**: Run a block of code multiple times.
    -   Iterate a set number of times: for loop
    -   Iterate until a condition is met: while loop
-   **Functions**:
    -   Associate a name and set of arguments with a complex series of commands.
-   We will focus on **conditional statements** this time.

## Review of logical variables and comparison operators

Logical variables only take two values: TRUE or FALSE.

-   Operators to use with conditionals: \>, \<, \>=, \<=, ==, !=, & (or &&), \| (or \|\|).

-   Let's create expressions that evaluate to TRUE or FALSE with each of the logical operators.

```{r}
# less than (or equal to)
5 <= 6
5 < 4

# greater than (or equal to)
6 > 5
5 >= 4

# equals/doesn't equal
"c" == "c"
TRUE != FALSE

# multiple - false
TRUE != FALSE & 1 == 2

# multiple - true
1 == 2 | 2 == 3 | 3 == 3
```

## Applying conditional operators to vectors

**any()**: tests whether any element of a vector is true.

-   Try getting a TRUE and FALSE below (in two different examples). Don't repeat the one I used.

```{r}
any(c("a","b","c") == "d")
# lets get a false 
cats = c("cats", "cats", "cats")
any(cats == "dogs")
# lets get a true
vec = 1:10
any(c(1:10) == 7)
```

**all()**: tests whether all elements of a vector are true.

-   Let's try getting a TRUE and FALSE below (in two different examples). Don't repeat the one I used.

```{r}
all(c(1,2,3,4,5) >=1)
# let's generate a true
cs = c("c","c","c")
all(cs == "c")
# let's generate a false
ts = c(TRUE,TRUE,FALSE)
all(ts == TRUE)
```

**%in%** is very similar to any(): tests whether an element is in a vector. We won't use this one often, but try getting the same answer as the ANY code chunk above.

```{r}

"dogs" %in% cats # false
"cats" %in% cats # true

7 %in% vec # true
11 %in% vec # false
```

Some practice questions

```{r}
test <- c(1,2,3,4)

# Check if there is any 5 in the vector above

# Check if all the elements of the vector test are positive

# Check if all the elements are smaller than or equal to 2 OR greater than or equal to 3
```

## Conditional syntax (if, else if, else)

#### Simple if...

if (condition is TRUE) { run the conditional code }

```{r}
Y = 24
if(is.numeric(Y)== TRUE){  
   print("Y is a NUMBER!")
}
```

if statements with object assignment:

```{r}
Y = 24
if(Y > 0){  
   x = sqrt(Y);print(x)
}
```

Other examples (one of the below is an error, fix it)

```{r}
T = 10
if(is.numeric(t)== TRUE){  
   print("Y is a NUMBER!")
}


value = "cats"
if(value/2 == 0){  
   x = sqrt(animal);print(x)
}
```

#### Conceptual if... else...

if (weather == sunny) {

go for a hike } else {

go to the museum

}

Let's look at one in R:

```{r}
Height = 59
if(Height < 66){  
   print("SHORT!")
} else if((Height >= 66) & (Height < 72)){ 
   print("MEDIUM!")
} else if(Height >= 72){  
   print("TALL!")
}
```

## Nested conditionals

Traffic light pseudocode:

if(light is green){

drive

} else if (light is yellow){

drive

} else if (intersection is still far) {

stop

} else if (light is red) {

stop

}

Write the converted traffic light example in the box below:

```{r}

```

## ifelse() statements

For simple conditionals or to perform vectorized operations, we can use the **ifelse()** function. ifelse() can evaluate entire vectors in one sweep.

Single values

```{r}
# categorizing a single value
x <- 1
ifelse(x > 0, "x is positive", "x is not positive")
```

Vectors

```{r}
# categorizing a vector
age <- c(17,16,20,18,22,18,15)
ifelse(age >= 18, "adult", "minor")
```

## Practice Exercises

1)  Create the following variables:

```{r}
g <- 9.8
x <- 12
y <- 4
bestCodingLanguages <- c("Python", "Matlab", "C++", "javaScript")
```

2)  Write an if statement to print "g is greater than 10" if g is greater than 10.

```{r}
if(g > 10){
  print("g is greater than 10")
}
# does not print anything because g is less than 10
```

3)  Write an if statement to print "x is divisible by y" if x is divisible by y. (Hint: The remainder of the division should be 0 - remember the %% operator.)

```{r}
if(x %% y == 0) {
  print("x is divisible by y")
}
```

4)  Write an if statement so that if "R" is in bestCodingLanguages, it prints "That's right!", and if not, it prints "It should be!".

```{r}
if("R" %in% bestCodingLanguages){
  print("That's Right!")
} else if (all(bestCodingLanguages != "R")){
  print("It should be.")
}

# can also just do else
if("R" %in% bestCodingLanguages){
  print("That's Right!")
} else{
  print("It should be.")
}

# can also use any 
if(any(bestCodingLanguages == "R")){
  print("That's Right!")
} else{
  print("It should be.")
}

# let's fix this
# to add an element, we can take the total number of elements + 1 so we don't overwrite anything
bestCodingLanguages[length(bestCodingLanguages)+1] <- "R"

# now let's check it 
if("R" %in% bestCodingLanguages){
  print("That's Right!")
} else if (all(bestCodingLanguages != "R")){
  print("It should be.")
}

# can also just do else
if("R" %in% bestCodingLanguages){
  print("That's Right!")
} else{
  print("It should be.")
}

# can also use any 
if(any(bestCodingLanguages == "R")){
  print("That's Right!")
} else{
  print("It should be.")
}

# NOTE: If you rerun the above code, it will say "That's right" for all of them because you have added that element. You need to reassign the objects in the initial code chunk to see "It should be."
```

5)  Re-write the code from 4) using the ifelse() function

```{r}
# reassign object so we can test
bestCodingLanguages <- c("Python", "Matlab", "C++", "javaScript")

ifelse(all(bestCodingLanguages != "R"), "It should be","That's Right") # prints "it should be" if all of the languages aren't R
ifelse(any(bestCodingLanguages == "R"), "That's Right","It should be") # prints "THat's right" if any of the languages are R
ifelse("R" %in% bestCodingLanguages, "That's Right","It should be") # prints "THat's right" if R is in best languages

# let's fix this
bestCodingLanguages[5] <- "R"
ifelse(all(bestCodingLanguages != "R"), "It should be","That's Right") # prints "it should be" if all of the languages aren't R
ifelse(any(bestCodingLanguages == "R"), "That's Right","It should be") # prints "THat's right" if any of the languages are R
ifelse("R" %in% bestCodingLanguages, "That's Right","It should be") # prints "THat's right" if R is in best languages
```

6)  Let's implement the traffic light example. Create the following dataframe showing the light color and distance (in feet) between the car and the light.

```{r}
df <- data.frame(color = c("green","green","yellow","yellow","red","red"),
                 distance = c(50,200,50,200,50,200))
df
```

Let's assume that the "distance of no return" (i.e. where the car has enough time to stop) is 125 feet. Add a column called "action" that says whether the car should "drive" or "stop" depending on the light color and distance.

a)  Do this using data frame indexing

```{r}
df[df$color == "green" | (df$color == "yellow" & df$distance <= 125), "action"] <- "drive" # this satisfies both drive conditions. Now let's do stop conditions
df[df$color == "red" | (df$color == "yellow" & df$distance > 125), "action"] <- "stop"
df
# this method is super ugly and clunky
```

b)  Do this using ifelse()

```{r}
df$action2 <- ifelse((df$color == "green" | (df$color == "yellow" & df$distance <= 125)), "drive","stop") # this is way easier 
```

## The end

That's it! Great job! :)
