---
title: "SOLUTIONS Lecture 4: Loops"
author: "PSYC417"
format: docx
editor: visual
---

# Loops Workbook

## Loops

Loops are one of the building blocks of algorithms. They allow us to
iterate through a segment of code repeatedly. This will be quite useful
to us when we run computational models, create functions, and in many
other situations. Loops may be confusing when you first learn about
them, but all you are doing is looping through normal bits of code. (The
trick is we will be indexing a lot!)

There are different types of loops. Specifically, we will focus on:

- *for* loops

- *nested* loops

- *while* loops.

### *for* loops

*for* loops (and loops in general) have a similar structure to
conditionals. You call the command (for or while), have an acceptable
"argument" in parentheses, and contain the body of body within curly
braces.

for(index in iteration){

Body of code.

}

for(i in 1:5){

print("HOLY MOLY")

}

#### **Examples of for loops**

Using a for loop, print the numbers between one and ten.

```{r}
for(i in 1:10){
  print(i)
}
```

Using a for loop, index an existing vector with the numbers one through
ten and print each element of the vector.

```{r}
numbers <- c(1:10)
for(i in 1:length(numbers)){
  print(numbers[i])
}
```

#### The paste command

The paste command is useful for printing strings or stitching together
values.

Paste "Meg is five years old." using the variables below. (This does not
require looping.)

```{r}
students <- c("Miles","Christian","Meg")
age <- 5

paste(students[3], "is",age,"years old.")
```

There is a space by default between pasted elements, but you can chance
it to *whatever* you want using the *sep=* argument and putting a value
in quotation marks.

```{r}
paste("What","in","the","world",sep = "I went to the store")
paste("What", "in", "the", "world", sep=".")
```

If you want there to be no spaces, then you use sep="" OR the paste0
function with no *sep* argument. This is especially useful for stitching
numbers together.

```{r}
as.numeric(paste(1,2,3,sep=""))
as.numeric(paste0(1,2,3))
```

#### Practice 1

1.) Create the following variables:

```{r}
student <- c("John", "Mary", "Alex", "Robin")
state <- c("Maryland", "New Mexico", "Kansas", "Iowa")
pronoun <- c("his (or he)","her (or she)","their (or they)","her (or she)")
gpa <- c(3.8, 3.5, 3.9, 3.8)
psychology_major <- c("no", "yes", "yes", "no")
df <- data.frame(student, state, pronoun, gpa, psychology_major)
df
```

2.) For each student, print the sentence "\[name\] is from \[state\] and
\[pronoun\] gpa is \[gpa\]"

Note: within a loop, paste() on its own will not print the output, you
need print(paste(...))

```{r}
for(i in 1:nrow(df)){ # we use 1:nrow(df) because each row corresponds to a student and we want a statement for each student. 
  print(paste(df$student[i],"is from", df$state[i],"and",df$pronoun[i],"gpa is",df$gpa[i])) # We must index the ith element since the for loop specifies i in... On the first run of the loop, i =1, so it will select the first student, first state, first pronoun, and first gpa, then the code ends so it sets i to 2, selects the second student, second state... and so on until i=4
}
```

3.) For each student, print the sentence "\[name\] is cool and
\[pronoun\] love(s) psychology." OR "\[name\] is/are NOT COOL and
\[pronoun\] do/does not love psychology." depending on whether they are
a psychology major or not. Remember: You have to print to view a paste
command in R.

```{r}
for(i in 1:nrow(df)){
  # We need a conditional to see whether they are a psychology major. Since there are two options, we can use an if and else statement. 
  if(df$psychology_major[i] == "yes"){ # will only run for psych majors
    print(paste(df$student[i],"is cool and", df$pronoun[i],"love(s) psychology.")) # grabs the name and pronoun of the ith student that is a psych major
  } else{ # this will account for situations where they are NOT a psych major
    print(paste(df$student[i],"is NOT cool and", df$pronoun[i],"do/does not love psychology.")) # print the student name and pronoun 
  }
}
```

4.) Add a column named "score" to the data frame. Score should be their
gpa multiplied by two, multiplied by whether they are a psychology major
(so you may need to convert this using a data type coercion...). Use a
for loop.

```{r}

for(i in 1:nrow(df)){
  df$score[i] <- df$gpa[i]*2*as.numeric(df$psychology_major[i]=="yes") # converts to 1 if psychology major, 0 if not since 1 is TRUE and 0 if FALSE.
}
```

Now do that without a loop.

```{r}
df$score_noloop <- df$gpa*2*as.numeric(df$psychology_major=="yes")
```

Which one is better?

The one without the loop is better. If you can perform a vectorized
operation in R, it will always perform better (i.e., faster) than
iterating with a loop. The time difference is negligible for small data,
but it's too bad of a programming habit to develop so don't iterate/loop
unless necessary.

### Nested Loops

You can place loops within each other and iterate through commands. Here
is an example of the logic of nested loops.

```{r}
for (i in 1:5){
   for(j in 1:5){
	   print(paste(i, "times", j, "is", i*j))
   }
}
```

#### Practice 2

1.) Let's print all the combinations of color, shade and transparency
from the following three vectors (e.g. "light opaque red", "dark
transparent blue"). How many combinations are there in total?

```{r}
shades <- c("light", "dark")
transparency <- c("transparent","opaque")
colors <- c("red","green","blue")

# i comment which braces belong to which loop when they're nested so I know what's happening. It can be hard sometimes if you appropriately indent things.
for(s in 1:length(shades)){ # start shades loop
  for(t in 1:length(transparency)){ # start transparency loop
    for(c in 1:length(colors)){ # start colors loop
      print(paste(shades[s],transparency[t],colors[c]))
    } # end colors loop
  } # end transparency loop
} # end shades loop
```

**How can we see how many conditions there are when looping like this?**

- We know that in terms of combinations, it's the number of
  possibilities for each category multiplied together.

```{r}
paste("There are",length(shades)*length(transparency)*length(colors),"combinations in total.")
```

### While Loops

While loops will iterate across code until a condition is false. (In
other words... while a condition is true.) While loops can be more
efficient than for loops since for loops have a fixed number of
iterations. If you are unsure how many iterations are required, or need
to stop the first X number of times that something happens, while loops
will be useful.

The structure of a while loop is similar, but we usually create some
kind of counter/index outside of a loop and increment that manually each
time. Additionally, you usually create a condition outside of the loop
that will be resolved in the while loop. **Note:** Do not forget to
iterate your counter or toggle an end to the while loop or it will run
forever!

```{r}
counter =1; condition = TRUE
while(condition == TRUE){
  if(counter > 5){
    print(counter)
    condition = FALSE
  }
  counter = counter + 1
}
```

#### Practice 3

Run the following code:

```{r}
df <- data.frame(id = c(1:20), 
                 age = c(20,22,23,18,18,19,19,21,20,19,19,20,22,22,21,21,23,18,17,20))
```

1.) Find the first 3 people who are younger than 20 and return their ID
in a vector called id_young

```{r}
count_young <- 0 # this is our counter of people younger than 20
id_young <- c() # this is where we will store their ID
i <- 1 # this is the counter to iterate through the rows of df
while(count_young < 3){ # once count_young = 3, this loop will end.
  if(df$age[i] < 20){ # we want to do things if someone is younger than 20, we don't care about anyone else
    id_young = c(id_young, df$id[i]) # combine id_young with the current df_id that is attached to a young person
    count_young = count_young + 1 # this will add one for each young person until we reach three.
  }
  i = i + 1 # we must do this because while loops do not manually adjust i for us...this will help us keep track of what row of the df we're in. DO NOT FORGET THIS OR THE CODE WILL NEVER STOP RUNNING!!!
}

# an equally valid way to do this, do whichever makes more sense to you
while (count_young < 3) {  
	if (df$age[i] < 20) {    
		count_young <- count_young + 1 # sets count_young to 1 for the first young person, adds 1 thereafter
		id_young[count_young] <- df$id[i] # indexes the count_youngth element of id_young. In other words, this will index the first element of id_young for the first young person, 2nd for the second young person, and third for the third young person. 
	}  
	i <- i + 1 
}
id_young
```

2.) What if you try to find the first 15 people (but there aren't
actually 15 people younger than 20 in your sample)?

```{r}
count_young <- 0 #this is our counter of people younger than 20
id_young <- c() #this is where we will store their ID
i <- 1 #this is the counter to iterate through the rows of df

# this is the same code as above but with while(count_young < 15). This will generate an error here but sometimes the code will run forever. Be careful! 
while(count_young < 15){ # once count_young = 3, this loop will end.
  if(df$age[i] < 20){ # we want to do things if someone is younger than 20, we don't care about anyone else
    id_young = c(id_young, df$id[i]) # combine id_young with the current df_id that is attached to a young person
    count_young = count_young + 1 # this will add one for each young person until we reach three.
  }
  i = i + 1 # we must do this because while loops do not manually adjust i for us...this will help us keep track of what row of the df we're in. DO NOT FORGET THIS OR THE CODE WILL NEVER STOP RUNNING!!!
}
```

While loops can be more efficient, but you have to be careful because
they can cripple your computing power if they run into an error. You
will have to hit the stop button in the terminal or stop in the job in
another (e.g., terminate your R session) to resume normal R operations.

## The end

That's it! Great job! :)
