---
title: "Lab 1: BUGS_BUGS_BUGS"
author: "HELP!"
format: docx
editor: visual
---

## War on bugs

From: confused_professor

To: programming_expert

Please help me with the below code. I need to submit this paper ASAP but
I'm just starting to learn R and don't know exactly what to do. I
appreciate it.

Best,\
CP

Here, I'm trying to create different logical types...

```{r}
logi = true
priNt(logi)
logi = logI + 1 # trying to add one to the logical value...
```

Now, I'm trying to work with that logical value some more.

```{r}
logi = logi*45LL
CLASS(LOGI) # what data type is it?

# convert the object to character data 
char_logi = is.string(logi)
```

```{r}
# I wanted 5 1s in an object
Nums1 == rep(1, 5).

# I want to mulitply those numbers by two
Nums2<-Nums1 * "2"
```

I have data I want to plot below. What is wrong?

```{r}
height = runif(100, 60, 100) weight = runif(100, 150, 200) 

# create the plot
plt <-- plot(Height, Weight); 

print the plot below, this is a comment
plt

# make sure this informative error message shows
print('The plot above is a scatterplot")
```

Do not delete any of the lines above, just fix them so the document
runs. Knit and submit on ELMS!
